Can dike grassland restoration profit from ideal seed mixture-substrate combinations?

### 2023-02-15 manuscript.Rmd
Update manuscript.Rmd

Sowing is a well established restoration technique to overcome dispersal limitation. Though, adapted seed mixtures to certain site conditions promise a more effective restoration. This is especially important if the restored vegetation has to protect constructions like dikes. An ideal seed mixture-substrate combination would enhance erosion protection and species-rich grasslands.

@@

persistences

@@

Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021]. For this reason, restoration ecology has to increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015]. Rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory are needed to guide evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Seeding in general is a way to overcome the dispersal filter [@myers2009; @orrock2023], but we focus on environmental and the biotic filtering which are the main assembly processes to be manipulated by the choice of the right seed mixture-substrate combination [@Torok.2017]. This means a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion of unsuited species. Modifying seed mixtures for certain site conditions came up with functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020] but it is not easy to put it into practice for novel ecosystems like urban grasslands [@Bauer.2022]. Substrates cannot only be modified for conservation purposes in urban areas, but also in quarries [@chenot-lescure2022] or dikes [@Liebrand.1996].

@@

To measure the success of restoration, we chose four indices. First, we observed the persistence of the sown species, which is a key interest of seeding as a restoration approach [@Wilsey.2021]. For this, it was surveyed if the sown species are present in the respective plot in following years. Second, we calculated the Favorable Conservation Status (FCS) which distinguishes characteristic diversity and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are target species for a habitat type of interest. Third, the successional distance to the reference was determined. This measure quantifies the distance of a plot to the average reference site from an ordination [@Rydgren.2019]. To complement the indices, the success of restoration or the difference from desired conditions needs to be evaluated [@Brudvig.2017b]. Therefore, the species composition of the experimental plots were compared with reference sites. We asked following question to evaluate the restoration approaches and success after four years:

@@

3.  How close is the vegetation to the reference state?

@@

We could identify a statistically clear positive effect of sand addition on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). Furthermore, sown hay meadows were clearly closer to their reference plots than sown dry grasslands were to their references. The posterior distributions which are shown in the interactions plots which separates exposition and survey year indicate for all three responses a strong positive development after one year with slow increase in following years (Figures \@ref(fig:persistence); \@ref(fig:fcs); \@ref(fig:recovery)) . Both expositions have similar trends but for all responses the values are clearly lower on the south exposition, e.g., persistence values are on average more than 46% higher on the north exposition (Figure \@ref(fig:persistence)). The effect of restoration treatments is in the interaction plots not clear. Hay meadow seed mixture led to a closer development to hay meadow references and almost reached on the north exposition the area of the standard deviation of reference sites (Figure \@ref(fig:recovery)). Though, the sand admixture had no clear effect on the development. Persistence of both seed mixture types seems to be slightly positive affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:persistence)).

@@

Our aim was to identify a ideal seed mixture-substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the establishment and persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions [Figure @ref(fig:traits); @Oberdorfer.1993; @Oberdorfer.1993b]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. Though, these two studies did not work with an experimental set up of different seed mixture-substrate combinations, but analyzed the result of habitat and biotic filtering after five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the sown species could establish earlier than others and preempted the available niches [@Fukami.2015].

@@

The vegetation developed in general in the right direction but was still distinct from positive references and seed mixtures after four years. It was rather similar to the negative reference of perennial ruderal vegetation. The gap between goal and restoration outcome was also shown for other seeding experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural references [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018].

@@

Our results suggest that adapted seed mixtures can increase restoration efficiency by seeding hay meadows on the north exposition and not on the south exposition of dikes. Furthermore, to reduce nutrient load by sand admixture has a clear positive effect, although the effect is small. The question remains if money is best invested in sand admixture or in other restoration measures. It seems not to be necessary to increase seeding density on dike slopes, which would reduce restoration costs. As well as the use of more deeper soil depth which would otherwise create costs for removal of topsoil. There was no perfect seed mixture-substrate combination, why we conclude that a variation of seed mixtures types and different soils along restoration sections would promote biodiversity better than a single solution [@bauer2023; @Holl.2022]. The strong and lasting negative effect of drought on the south exposition might require a resowing. To close the recovery debt, the adaption of the management might be promising since it is a relevant factor aside of restoration approach and site characteristics [@Grman.2013; @tölgyesi2021]. For example, starting sheep grazing on the restored sites will modify the disturbance regime and improve dispersal. All in all, our results support the finding that dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. Though, the recovery debt highlights that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020; @Buisson.2022].

>>>>

1.  Sowing is a well established restoration technique to overcome dispersal limitation. Though, adapted seed mixtures to certain site conditions promise a more effective restoration. This is especially important if the restored vegetation has to protect constructions like dikes. An ideal seed mixture-substrate combination would concurrently enhance erosion protection and species-rich grasslands by a fast and successful establishment of target species.
2.  An experiment with 288 plots was conducted on both expositions of a dike and vegetation was surveyed 2018-2021. We tested different sand admixtures, substrate depths, seed densities and seed mixture types (hay meadows, dry calcareous grasslands), and measured the persistence of sown species, the Favorable Conservation Status (FCS) and the successional distance to reference sites as calculated from an ordination.
3.  Sand admixture had a significant but slightly positive effect on all indices. However, substrate depth, seed density and seed mixture type had no significant effect on persistence or FCS.
4.  On the north exposition, hay meadow-seed mixtures developed closer to its reference than dry grasslands. Overall, the vegetation developed in the desired direction but a recovery debt remained after four years. Plots still exhibited similarities to negative references from ruderal vegetation.
5.  The south exposition was significantly worse than the north exposition which might be due to the severe drought during establishment and the addition and removal of a geotextile. This negative effect remained over the whole observation period.
6.  Synthesis and applications: Adapted seed mixtures can make restorations more efficient. Though, varying seed mixture-substrate combinations might foster biodiversity more than a desired 'ideal' combination. Furthermore, additional efforts in management or resowing due to severe legacy effects by drought might be necessary to reduce the recovery debt.

@@

persistence

@@

Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts [@Freitag.2021]. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021]. For this reason, restoration ecology has to increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015]. Rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory are needed to guide evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Sowing in general is a way to overcome the dispersal filter [@myers2009; @orrock2023], but we focus on environmental and the biotic filtering which are the main assembly processes to be manipulated by the choice of the right seed mixture-substrate combination [@Torok.2017]. This means a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion of unsuited species. Modifying seed mixtures for certain site conditions came up with functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020] but it is not easy to put it into practice for novel ecosystems like urban grasslands [@Bauer.2022]. Substrates cannot only be modified for conservation purposes in urban areas, but also in quarries [@chenot-lescure2022] or dikes [@Liebrand.1996].

@@

To measure the success of restoration, we chose four indices. First, we observed the persistence of the sown species, which is a key interest of sowing as a restoration approach [@Wilsey.2021]. For this, it was surveyed if the sown species are present in the respective plot in following years. Second, we calculated the Favorable Conservation Status (FCS) which distinguishes characteristic diversity and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are target species for a habitat type of interest. Third, the successional distance to the reference was determined. This measure quantifies the distance of a plot to the average reference site from an ordination [@Rydgren.2019]. To complement the indices, the success of restoration or the difference from desired conditions needs to be evaluated [@Brudvig.2017b]. Therefore, the species composition of the experimental plots were compared with reference sites. We asked following question to evaluate the restoration approaches and success after four years:

@@

3.  How close is the established vegetation to the reference state?

@@

We could identify a statistically clear positive effect of sand addition on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). Furthermore, sown hay meadows were clearly closer to their reference plots than sown dry grasslands were to their references. The posterior distributions which are shown in the interactions plots which separates exposition and survey year indicate for all three responses a strong positive development after one year with slow increase in following years (Figures \@ref(fig:persistence); \@ref(fig:fcs); \@ref(fig:recovery)) . Both expositions have similar trends but for all responses the values are clearly lower on the south exposition, e.g., persistence values are on average more than 46% higher on the north exposition (Figure \@ref(fig:persistence)). The effect of restoration treatments is in the interaction plots not clear. Hay meadow-seed mixture led to a closer development to hay meadow references and almost reached on the north exposition the area of the standard deviation of reference sites (Figure \@ref(fig:recovery)). Though, the sand admixture had no clear effect on the development. Persistence of both seed mixture types seems to be slightly positive affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:persistence)).

@@

Our aim was to identify a ideal seed mixture-substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the establishment and persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions [Figure @ref(fig:traits); @Oberdorfer.1993; @Oberdorfer.1993b]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. Though, these two studies did not work with an experimental set up of different seed mixture-substrate combinations, but analyzed the result of habitat and biotic filtering after five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the sown species could establish earlier than others and preempted the available niches [@Fukami.2015].

@@

The vegetation developed in general in the right direction but was still distinct from positive references and seed mixtures after four years. It was rather similar to the negative reference of perennial ruderal vegetation. The gap between goal and restoration outcome was also shown for other sowing experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural references [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018].

@@

Our results suggest that adapted seed mixtures can increase restoration efficiency by sowing hay meadows on the north exposition and not on the south exposition of dikes. Furthermore, to reduce nutrient load by sand admixture has a clear positive effect, although the effect is small. The question remains if money is best invested in sand admixture or in other restoration measures. It seems not to be necessary to increase seeding density on dike slopes, which would reduce restoration costs. As well as the use of more deeper soil depth which would otherwise create costs for removal of topsoil. There was no perfect seed mixture-substrate combination, why we conclude that a variation of seed mixtures types and different soils along restoration sections would promote biodiversity better than a single solution [@bauer2023; @Holl.2022]. The strong and lasting negative effect of drought on the south exposition might require a resowing. To close the recovery debt, the adaption of the management might be promising since it is a relevant factor aside of restoration approach and site characteristics [@Grman.2013; @tölgyesi2021]. For example, starting sheep grazing on the restored sites will modify the disturbance regime and improve dispersal. All in all, our results support the finding that dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. Though, the recovery debt highlights that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020; @Buisson.2022].
